Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands

### 2023-03-01 manuscript.Rmd
Update manuscript.Rmd

Grasslands can support an exceedingly high biodiversity and they provide several ecosystem services [@Dengler.2014; @Bardgett.2021]. However, they are globally endangered [@Bardgett.2021] for example by misguided afforestation [@dudley2020], and in Europe, calcareous grasslands and hay meadows are red-listed habitats [Category 3, 'vulnerable', @Janssen.2016]. Restoration is seen as a key factor to sustain biodiversity and ecosystem services [@COP.2014; @UnitedNations.2019b], and sowing is a well-established approach to establish species-rich grasslands [@Kiehl.2010]. Sowing high-diversity mixtures of local provenance produced by specialized companies is a promising way to scale up restoration efforts [@Freitag.2021], and to overcome the dispersal filters [@myers2009; @orrock2023]. However, there are still open questions about adjusting the seed mixtures to specific site conditions and future climate conditions [@Torok.2021].

Restoration ecology can increase the predictability of restoration approaches [@Suding.2011; @Mouquet.2015] by using rigorous, repeatable, and transparent experiments based on advanced theory, which will finally lead to evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration method are key predictors of the development of vegetation after sowing [@Brudvig.2017b], and habitat and biotic filtering are the main assembly factors which can be manipulated by the choice of the seed--substrate combination [@Torok.2017]. This means a close adaptation of the substrate to the niche of the target species or of the seed mixtures to the characteristics of the chosen substrate. Suitable substrates should reduce habitat filtering, while the seed mixture should minimize the competitive exclusion of unsuitable species. Modifying seed mixtures to match specific site conditions could be based on functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020], although this is not easy to implement [@merchant2022; @Bauer.2022]. This challenge is particularly interesting on artificial substrates that are used in urban areas [@Bauer.2022], quarries [@chenot-lescure2022], or on dikes [@Liebrand.1996].

@@

The aim of this study is to identify the best combination of seed mixtures and substrates for vital and species-rich grasslands on north- and south-exposed dike slopes. An experiment was set up to test agricultural soil with different amounts of sand admixture, substrate depths, seed densities, and seed mixture types. We expected a better development of calcareous grassland in the south exposition with shallow sandy substrates, and mesic hay meadows in north exposition on less sandy deep substrate. Furthermore, higher seed densities were recommended for slopes [@KleberLerchbaumer.2017], albeit without experimental evidence.

To measure the success of restoration, we chose three indices. First, we observed the persistence of the sown species, which is a key objective of this restoration measure [@Wilsey.2021]. For this purpose, presence of the sown species was monitored for three consecutive years. Second, we calculated the Favourable Conservation Status (FCS) which distinguishes characteristic diversity and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are associated with the habitat type of interest, whereas derived species not. Third, the recovery was determined by measuring the successional distance to the reference grasslands [@Rydgren.2019]. To complement these indices, the success of restoration or the difference from desired conditions was evaluated [cf. @Brudvig.2017b]. This was reached by comparing the species composition of the experimental plots with reference sites. We asked the following question to evaluate the restoration approaches and the success after four years:

1.  Mesic vs dry grassland mixtures are favoured by site conditions on northern and southern dike slopes, respectively, compared to their reference sites.

2.  Nutrient reduction (by sand addition) and lower soil depth increase persistence of sown species and fosters target species in comparison to non-target species especially for dry grassland mixtures. Hay meadow mixtures can cope with no sand admixture and larger soil depths.

3.  Higher seed density increases generally persistence of sown species and suppresses non-target species.

>>>>

Grasslands can support an exceedingly high biodiversity and they provide several ecosystem services [@Dengler.2014; @Bardgett.2021]. However, they are globally endangered [@Bardgett.2021], and in Europe, calcareous grasslands and hay meadows are red-listed habitats [Category 3, 'vulnerable', @Janssen.2016]. Restoration is seen as a key factor to sustain biodiversity and ecosystem services [@COP.2014; @UnitedNations.2019b], and sowing is a well-established approach to establish species-rich grasslands [@Kiehl.2010]. Sowing high-diversity mixtures of local provenance produced by specialized companies is a promising way to scale up restoration efforts [@Freitag.2021], and to overcome the dispersal filters [@myers2009; @orrock2023]. However, there are still open questions about adjusting seed mixtures to specific site conditions and future climate conditions [@Torok.2021].

Restoration ecology can increase the predictability of restoration approaches [@Suding.2011; @Mouquet.2015] by using rigorous, repeatable, and transparent experiments based on advanced theory, which will finally strengthen evidence-based restoration [@Cooke.2018; @Wainwright.2018]. Local site conditions and the restoration method are key predictors for vegetation development after sowing [@Brudvig.2017b], and habitat and biotic filtering are the main assembly factors which can be manipulated by the choice of seed--substrate combinations [@Torok.2017]. This means a close adaptation of the substrate to the niche of the target species or of the seed mixtures to the characteristics of the chosen substrate. Suitable substrates should reduce habitat filtering, while the seed mixture should minimize the competitive exclusion of unsuitable species. Modifying seed mixtures to match specific site conditions could be based on functional plant traits [@Funk.2008; @Laughlin.2014; @Balazs.2020], although this is not easy to implement [@merchant2022; @Bauer.2022]. This challenge is particularly interesting on artificial substrates that are used in urban areas [@Bauer.2022], quarries [@chenot-lescure2022], or on dikes [@Liebrand.1996].

@@

The aim of this study is to identify suitable combinations of seed mixtures and substrates for vital and species-rich grasslands on north- and south-exposed dike slopes. An experiment was set up to test agricultural soil with different amounts of sand admixture, substrate depths, seed densities, and seed mixture types. We expected a better development of calcareous grassland in the south exposition with shallow and sandy substrates, and of mesic hay meadows in north exposition on less sandy and deep substrates. Furthermore for slopes, higher seed densities are recommended [@KleberLerchbaumer.2017], albeit without experimental evidence.

To measure the success of restoration, we chose three indices. First, we observed the persistence of the sown species, which is a key objective of this restoration measure [@Wilsey.2021]. For this purpose, the presence of the sown species was monitored for three consecutive years. Second, we calculated the Favourable Conservation Status (FCS) which distinguishes characteristic and derived diversity [@Helm.2015]. Characteristic diversity is created by established species which belong to the habitat-specific species pool. In other words, these species are associated with the habitat type of interest, whereas derived species not. Third, the recovery was determined by measuring the successional distance to the reference grasslands [@Rydgren.2019]. To complement these indices, the success of restoration or the difference from desired conditions was evaluated [cf. @Brudvig.2017b]. This was reached by comparing the species composition of the experimental plots with reference sites. We tested following hypotheses after four years of observation:

1.  Establishment of mesic vs dry grassland mixtures is facilitated by site conditions on northern and southern dike slopes, respectively.

2.  Nutrient reduction (by sand addition) and shallow soil increase establishment of dry grassland compared to mesic mixtures.

3.  Reduced soil resources benefit especially target species in comparison to non-target species(, thus increasing conservation value).

4.  Higher seed density increases generally establishment of sown plants and suppresses non-target species.
